Subsidiary Recursion in Coq

### 2022-01-15 paper/paper.tex
polishing, shrinking
Now just around 16 pages, still need conclusion and revised related work.

Central to interactive theorem provers like Coq, Agda, Isabelle/HOL,
Lean and others are terminating recursive functions over user-declared
inductive datatypes~\cite{agda,coq,isabelle-hol,lean}.  Termination is
usually enforced by a syntactic check for structural decrease.  This
structural termination is sufficient for many basic functions.  For
example, the well-known \texttt{span} function from Haskell's standard
library (\verb|Data.List|) takes a list and returns a pair of the
maximal prefix satisfying a given predicate \verb|p|, and the
remaining suffix:
\begin{verbatim}
span :: (a -> Bool) -> [a] -> ([a],[a])
span _ []    =  ([], [])
span p (x:xs) = if p x
                then let (ys,zs) = span p xs in (x:ys,zs)
                else ([],x:xs)
\end{verbatim}
\noindent The sole recursive call is \verb|span p xs|, and it occurs
in a clause where the input list is of the form \verb|x:xs|.  So the
input to the recursive call is a subdatum of the input, and hence this
definition is structurally decreasing.  In the appropriate syntax, it
can be accepted without additional effort by all the mentioned
provers.

This paper is about a more expressive form of terminating recursion,
called \textbf{subsidiary recursion}.  While performing an outer
recursion on some input \verb|x|, one may initiate an inner recursion
on \verb|x| (or possibly some of its subdata), preserving the
possibility of further invocations of the outer recursive function.
Let us see a simple example.  The function \verb|wordsBy| (from
\verb|Data.List.Extra|) breaks a list into its maximal sublists whose
elements do not satisfy a predicate \verb|p|.  For example,
\verb|wordsBy isSpace " good day "| returns \verb|["good","day"]|; so
\verb|wordsBy isSpace| has the same behavior as \verb|words| (from
\verb|Data.List|).  Code is in Figure~\ref{fig:wordsBy}.  The first
recursive call, \verb|wordsBy p tl|, is structural.  But in the
second, we invoke \verb|wordsBy p| on a value obtained from another
recursion, namely \verb|span|.  This is not allowed under structural
termination, but will be permitted by subsidiary recursion as derived
below.

\begin{figure}
\begin{verbatim}
wordsBy :: (a -> Bool) -> [a] -> [[a]]
wordsBy p [] = []
wordsBy p (hd:tl) =
  if p hd
  then wordsBy p tl 
  else let (w,z) = span (not . p) tl in
        (hd:w) : wordsBy p z
\end{verbatim}
\caption{Haskell code for \texttt{wordsBy}, demonstrating subsidiary recursion}
\label{fig:wordsBy}
\end{figure}

@@

This paper presents a functor-generic derivation of terminating
subsidiary recursion and induction in Coq.  We should emphasize that
this is a derivation of this recursion scheme within the type theory
of Coq.  No axioms or other modifications to Coq of any kind are
required. Based on this derivation, we present several example
functions like \verb|wordsBy|, and prove theorems about them.  A nice
example is a definition of run-length encoding using \verb|span| as a
subsidiary recursion, where we prove that encoding and then decoding
returns the original list.  Our approach applies to the standard
datatypes in the Coq library, and does not require switching libraries
or datatype definitions.

An important technical novelty of our approach is a derivation of a
weakened form of positive-recursive type in Coq.  Coq (Agda, and Lean)
restrict datatypes $D$ to be strictly positive: in the type for any
constructor of $D$, $D$ cannot occur to the left of any arrows.  Our
derivation needs to use positive-recursive types, where $D$ may occur
to the left of an even number (only) of arrows.  Coq requires strict
positivity because in the presence of other features of Coq's theory,
full positive-recursive types lead to a paradox~\cite{coquand88}.  We
present a way to derive a weakened form of positive-recursive type
that is sufficient for our examples (Section~\ref{sec:mu}).  The
weakening is to require only that $F\ \mu$ is a retract of $\mu$,
where $\mu$ is the recursive type and $F\ \mu$ its one-step unfolding.
Usually these types are isomorphic.  Hence, we dub these
\textbf{retractive-positive} recursive types.  This weakening has the
negative consequence of leading to a form of noncanonicity, but we
will see how to work around this.  Our definition of
retractive-positive recursive types makes essential use of impredicate
quantification, and hence cannot be soundly recapitulated in a
predicative theory like Agda's.

@@

We will use Coq inductive types for the signature functors \verb|F| of
various datatypes, thus enabling recursions to use Coq's
pattern-matching on the subdata structure \verb|d|.  So the style of
coding against this interface retains a similar feel to structural
recursions.  Unlike with structural termination, though, the interface
here is type-based and hence compositional.  As we will see, it
supports nested and higher-order recursions.

As in previous work, we dub this interface a \emph{recursion
universe}~\cite{stump20}.  As in other domains using the term
``universe'', we have an entity (here, \verb|R|) from which one cannot
escape by using the available operations (for other cases:
the ordinal $\epsilon_0$ and $\omega^-$, the physical universe and traveling at
the speed of light).  Staying in the recursion universe is good,
because we may recurse (via \verb|eval|) on any value of type
\verb|R|.

Some points must still be explained, particularly why \verb|X| has type \verb|Set -> Set|,
and the definition of \verb|FoldT|.  Let us see these and other details next.

@@

  This file specializes the development in \verb|Subrec.v| to the case
  of lists (parametrized by the type \verb|A| of elements).  In
  general, to use our development to get subsidiary recursion over
  some datatype, one will have a similar ``shim'' file.  For space
  reasons, we just give the example of lists.  The file
  defines the signature functor \verb|ListF|, shown in
  Figure~\ref{fig:listf}.  We then define \verb|List| to be
  \verb|Subrec|, with the instantiation of \verb|F| to \verb|ListF A|.
  This type \verb|List| is not to be confused with the type
  \verb|list| of lists in Coq's standard library.  As noted
  previously, our development is meant to be used in extension of
  existing inductive datatypes, not replacing them.  The figure also
  shows constructors \verb|mkNil| and \verb|mkCons| for \verb|List|,
  and types for conversion functions between \verb|List| and
  \verb|list| (see Section~\ref{sec:deriv} for the code).

@@

\verb|Subrec.v| also defines the notion of algebra that is used for
writing recursions.  The central definitions are in
Figure~\ref{fig:algf}.  \verb|KAlg| is the kind for the
type-constructor for algebras, as we see in the definition of
\verb|Alg|.  This type-constructor \verb|Alg| is a fixed-point of the
type \verb|AlgF|.  The fixed-point is taken using \verb|MuAlg|
(Section~\ref{sec:mu}), which implements our retractive-positive
recursive types at kind \verb|KAlg|.  Using \verb|Alg| will require
that \verb|AlgF| only uses its parameter \verb|Alg| positively.  We
will confirm this shortly.

@@

The type \verb|FoldT Alg C| is the type for fold functions which apply
algebras of type \verb|Alg| to data of type \verb|C|, which we have
already dubbed the \emph{anchor type} of the recursion.  At the top
level of code, the anchor type would just be \verb|List| (for
example).  When one initiates a subsidiary recursion, though, the
anchor type will instead by the abstract type \verb|R| for the outer recursion.

The variable \verb|Alg| occurs only positively (but not strictly
positively) in \verb|AlgF|, because it occurs negatively in
\verb|FoldT Alg R| which occurs negatively in \verb|AlgF Alg C X|.  So
we can indeed take a fixed-point of \verb|AlgF| to define the constant
\verb|Alg|.

@@

Given a predicate \verb|p : A -> bool|, and a value of type
\verb|List A|, we would like to compute a pair of type
\verb|list A * List A|, where the first component is the maximal
prefix whose elements satisfy \verb|p|, and the second is the
remaining suffix.  This is the typing for a top-level recursion.  More
generally, though, given an anchor type \verb|R : Set| along with a
fold function for that anchor type (i.e., of type \verb|FoldT (Alg (ListF A)) R|),
we would like to map an input list of type \verb|R| to a pair of
type \verb|list A * R|.  The first component of this pair is
going to be built up from scratch, and so cannot have type \verb|R|.
But the second component will be a subdatum of the input list,
and so can still have type \verb|R|.  This will enable outer recursions
to continue on that component.  So we want: 
\begin{verbatim}
Definition spanr{R : Set}(fo:FoldT (Alg (ListF A)) R)
                (p : A -> bool)(xs : R) : list A * R.
\end{verbatim}
\noindent From this we can also define the top-level recursion, by
supplying \verb|fold (ListF A)|, which is the function for folding an algebra
over a list (Figure~\ref{fig:algf}), for the argument \verb|fo| of \verb|spanr|:
\begin{verbatim}
Definition span(p : A -> bool)(xs : List A) : list A * List A
  := spanr (fold (ListF A)) p xs.
\end{verbatim}

Before we define \verb|spanr|, we must resolve a small problem.
If the first element of the input list
\verb|xs| to \verb|span| does not satisfy \verb|p|, then \verb|span|
should return \verb|([], xs)|.  But when recursing on \verb|xs|, we
will see it only in the form of a subdata structure of type
\verb|F R|.  We will not be able to return it from our recursion at
type \verb|R|, and hence we would not be able to return \verb|([],xs)|
as desired.  To work around this, we will have our recursion return a value
of type \verb|SpanF R|:
\begin{verbatim}
Inductive SpanF(X : Set) : Set :=
  SpanNoMatch : SpanF X
| SpanSomeMatch : list A -> X -> SpanF X.
\end{verbatim}
\noindent The idea is that the recursion will signal if it is in the
one tricky case where \verb|p| does not match the first element, by
returning \verb|SpanNoMatch|.  Otherwise, it will be able to return,
via \verb|SpanSomeMatch|, a prefix and the suffix at type
\verb|R|. The prefix will be nonempty, and hence the suffix will be at
most the tail of \verb|xs|.  This tail is available to the algebra
in the subdata structure of type \verb|F R|.

Figure~\ref{fig:span} gives the algebra \verb|SpanAlg| for computing
\verb|span|, and the code for \verb|spanr|.  We elide the proof
\verb|SpanFunctor| that \verb|SpanF| is indeed a \verb|Functor|,
and make \verb|X| implicit in the constructors of \verb|SpanF|.
The type of \verb|SpanAlg p C| is
\begin{verbatim}
Alg (ListF A) C SpanF
\end{verbatim}
This states that we are defining an algebra (\verb|Alg|) for the
\verb|ListF A| functor, with anchor type \verb|C| and carrier
\verb|SpanF|.  \verb|SpanF| has type \verb|Set -> Set|, as required
for the carriers of our algebras. The definition of \verb|SpanAlg| is
actually parametrized by \verb|C|, which is good, as it means we can
use \verb|SpanAlg| for top-level or subsidiary recursions.

Let us continue through the code for \verb|SpanAlg|
(Figure~\ref{fig:span}).  We use \verb|rollAlg| to create an algebra
from something whose type is an application of \verb|AlgF|.  This
takes in all the components of the recursion universe: the abstract
type \verb|R|, the \verb|reveal| function (not needed in this case),
the fold function (\verb|fo|) for any subsidiary recursions (also not
needed here), a function we choose to name \verb|span| for making
recursive calls, and finally \verb|xs : ListF A R|.  The algebra
pattern-matches on this \verb|xs|.  In the cases where it is empty or
where its head (\verb|hd|) does not satisfy \verb|p|, we return
\verb|SpanNoMatch|.  This signals to the caller that we really wished
to return \verb|([],xs)|, but could not because we do not have
\verb|xs| at type \verb|R|.  If the head does satisfy \verb|p|, then
we recurse on the tail (\verb|tl : R|) by calling the provided \verb|span : R -> SpanF R|.
If \verb|span tl|
returns \verb|SpanNoMatch|, that means that we should make \verb|tl|
the suffix in the pair we return (via \verb|SpanSomeMatch|).  Happily,
we have \verb|tl : R| here, so we can do this.  In either case (for
return value of \verb|span tl|), we add the head to the front of the
prefix.  We define \verb|spanhr| to invoke the fold function it is
given, on the algebra (\verb|SpanAlg|).

The final twist is now in the definition of \verb|spanr|.  We call
\verb|spanhr| on the input \verb|xs : R|.  If \verb|spanhr| returns
\verb|SpanNoMatch|, then we are supposed to return \verb|([],xs)|,
which we can do here, because we have \verb|xs : R|.  It was only
inside the algebra that we lost the information that the subdata
structure of type \verb|F R| is derived from a value of type \verb|R|.
If \verb|spanhr| returns \verb|SpanSomeMatch|, then the return value
gives us the nonempty prefix (\verb|l|) and the suffix (\verb|r|), which we then return.

\begin{figure}
\begin{verbatim}
Definition SpanAlg(p : A -> bool)(C : Set)
  : Alg (ListF A) C SpanF :=
  rollAlg (fun R reveal fo span xs => 
     match xs with
         Nil => SpanNoMatch 
       | Cons hd tl =>
          if p hd then
            match (span tl) with
              SpanNoMatch => SpanSomeMatch [hd] tl
            | SpanSomeMatch l r => SpanSomeMatch (hd::l) r
            end
          else
            SpanNoMatch 
       end).

Definition spanhr{R : Set}(fo:FoldT (Alg (ListF A)) R)
                 (p : A -> bool)(xs : R) : SpanF R :=
  fo SpanF SpanFunctor (SpanAlg p R) xs.

Definition spanr{R : Set}(fo:FoldT (Alg (ListF A)) R)
                (p : A -> bool)(xs : R) : list A * R
  := match spanhr fo p xs with
       SpanNoMatch => ([],xs)
     | SpanSomeMatch l r => (l,r)
     end.
\end{verbatim}
\caption{The algebra \texttt{SpanAlg} for the \texttt{span} function, and some functions based on it}
\label{fig:span}
\end{figure}

We can easily define \verb|break|, in Figure~\ref{fig:break}.  The
function \verb|breakr| is a version of \verb|break| that can be used
for subsidiary recursion, similarly to \verb|spanr| for \verb|span|.
Such a function always takes in a fold function (\verb|fo|) with
anchor type \verb|R|, which then is used to fold the algebra in
question.

\begin{figure}
\begin{verbatim}
Definition breakr{R : Set}(fo:FoldT (Alg (ListF A)) R)
                 (p : A -> bool)(xs : R) : list A * R :=
  spanr fo (fun x => negb (p x)) xs.

Definition break(p : A -> bool)(xs : List A) : list A * List A :=
  breakr (fold (ListF A)) p xs.
\end{verbatim}
\caption{The \texttt{break} function and its more flexible version, \texttt{breakr}, defined in terms of \texttt{spanr} (Figure~\ref{fig:span})}
\label{fig:break}
\end{figure}

@@

Let us now see how to write \verb|wordsBy|, our example function from
Section~\ref{sec:intro}, using \verb|spanr| as a subsidiary recursion.
The code is in Figure~\ref{fig:wordsby}, assuming a type
\verb|A : Set|.  The setup is similar to that for \verb|span|.  We first define
an algebra \verb|WordsBy|, parametrized by anchor type \verb|C| (and also
the predicate \verb|p|), of type
\begin{verbatim}
Alg (ListF A) C (Const (list (list A)))
\end{verbatim}
\noindent This says that \verb|WordsBy p C| is an algebra (\verb|Alg|)
for the \verb|ListF A| functor, with anchor type \verb|C|, and carrier
\verb|Const (list (list A))|.  \verb|Const| is the combinator for
creating the object part of constant functors; \verb|FunConst| creates
the morphism part (i.e., the \verb|fmap| function).  We use it
\verb|Const| here and in other examples where the return type of the
algebra will not depend on its abstract type \verb|R|.  Here, we are
constructing from scratch a list of lists, so it will not be legal to
recurse on the list itself, or its (list) elements.  Intead, we just
use the \verb|list| type of Coq's standard library.

@@

Here we see the benefit of our approach.  From Figure~\ref{fig:break},
the return type of \verb|breakr| is \verb|list A * R|, where \verb|R|
is the anchor type of the provided fold function \verb|fo|.  And
\verb|fo| has type \verb|FoldT (ListF A) Alg R|, from the definition
of \verb|AlgF| in Figure~\ref{fig:algf} (instantiating the functor
with \verb|ListF A|).  This means that from the invocation of
\verb|breakr|, we get \verb|w : list A| and \verb|z : R|.  And so we
can indeed apply \verb|wordsBy : R -> list (list A)| to
\verb|z| to recurse.

@@

The Haskell library \verb|Data.List.Extra| has a function \verb|repeatedly|, which
is defined essentially as follows; I have attempted a more informative name:
\begin{verbatim}
mapThrough :: (a -> [a] -> (b, [a])) -> [a] -> [b]
mapThrough f [] = []
mapThrough f (a:as) = b : mapThrough f as'
    where (b, as') = f a as
\end{verbatim}
\noindent The idea is that the function is like the standard \verb|map| function
on lists, except that here, the function \verb|f| that we are mapping (or ``mapping
through'') takes in not just the current element \verb|a|, but also the tail \verb|as|.
It then returns the value \verb|b| to include in the output list, and whatever other
list it wishes, upon which \verb|mapThrough| will recurse.

We can write this combinator using our infrastructure for subsidiary recursion.
For this to work, we need to supply the mapped function with the fold function
for \verb|mapThrough|'s recursion.  This is so that the mapped function can
initiate a subsidiary recursion, returning a value in the abstract type \verb|R|
of \verb|mapThrough|'s recursion.  So the type we will use for mapped functions
is:
\begin{verbatim}
Definition mappedT(A B : Set) : Set :=
  forall(R : Set)(fo:FoldT (Alg (ListF A)) R), A -> R -> B * R.
\end{verbatim}
\noindent This type is more informative than the Haskell type,
since it shows that the second component of the returned value
must have type \verb|R|, and hence must be (hereditarily) a tail
of the input of type \verb|R|.

Given this definition, the code for \verb|mapThrough| and
\verb|mapThroughr| is in Figure~\ref{fig:mapthrough}.  The code for
\verb|MapThroughAlg| is very similar (discounting syntax) to the
Haskell code above.  Here, though, when we call \verb|f|, we must
supply the abstract type \verb|R| and fold function \verb|fo|.  Then,
from the definition of \verb|mappedT|, we have that \verb|b : B| and
\verb|c : R|.  So we may indeed invoke \verb|mapThrough : R -> List B|
on \verb|c|.  Note that as we are building up a new list from scratch
(rather than just extracting some tail of the input list), we just
return \verb|list B|; we cannot perform further subsidiary recursion
on the output.

@@

Definition mapThroughr{R : Set}(fo:FoldT (Alg (ListF A)) R)
                      {B : Set}(f:mappedT A B) : R -> list B :=
  fo (Const (list B)) (FunConst (list B)) (MapThroughAlg f R).

>>>>

Central to interactive theorem provers like Coq, Agda, Isabelle/HOL,
Lean and others are terminating recursive functions over user-declared
inductive datatypes~\cite{agda,coq,isabelle-hol,lean}.  Termination is
usually enforced by a syntactic check for structural decrease, which
is sufficient for many basic functions.  For example, the
\texttt{span} function from Haskell's prelude (\verb|Data.List|) takes
a list and returns a pair of the maximal prefix whose elements satisfy
a given predicate \verb|p|, and the remaining suffix:
\begin{verbatim}
span :: (a -> Bool) -> [a] -> ([a],[a])
span _ []    =  ([], [])
span p (x:xs) = if p x
                then let (ys,zs) = span p xs in (x:ys,zs)
                else ([],x:xs)
\end{verbatim}
\noindent The sole recursive call is \verb|span p xs|, and it occurs
in a clause where the input list is of the form \verb|x:xs|.  Hence it
is structurally decreasing.  In the appropriate syntax, this
definition can be accepted without additional effort by all the
mentioned provers.

This paper is about a more expressive form of terminating recursion,
called \textbf{subsidiary recursion}.  While performing an outer
recursion on some input \verb|x|, one may initiate an inner recursion
on \verb|x| (or possibly some of its subdata), preserving the
possibility of further invocations of the outer recursive function.
Let us see a simple example.  The function \verb|wordsBy|
(\verb|Data.List.Extra|) breaks a list into its maximal sublists whose
elements do not satisfy a predicate \verb|p|.  For example,
\verb|wordsBy isSpace " good day "| returns \verb|["good","day"]|.
Code is in Figure~\ref{fig:wordsBy}.  Recall that
\verb|break p| is equivalent to \verb|span (not . p)|. The first recursive call,
\verb|wordsBy p tl|, is structural.  But in the second, we invoke
\verb|wordsBy p| on a value obtained from another recursion, namely
\verb|span|.  This is not allowed under structural termination, but
will be permitted by subsidiary recursion.

\begin{figure}
\begin{verbatim}
wordsBy :: (a -> Bool) -> [a] -> [[a]]
wordsBy p [] = []
wordsBy p (hd:tl) =
  if p hd
  then wordsBy p tl 
  else let (w,z) = break p tl in
        (hd:w) : wordsBy p z
\end{verbatim}
\caption{Haskell code for \texttt{wordsBy}, demonstrating subsidiary recursion}
\label{fig:wordsBy}
\end{figure}

@@

This paper presents a functor-generic derivation of terminating
subsidiary recursion and induction in Coq.  We should emphasize that
this is a derivation within the type theory of Coq, and requires no
axioms or other modifications to Coq. Using this derivation, we
present several example functions like \verb|wordsBy|, and prove
theorems about them.  A nice example is a definition of run-length
encoding using \verb|span| as a subsidiary recursion, where we prove
that encoding and then decoding returns the original list.  Our
approach applies to the standard datatypes in the Coq library, and
does not require switching libraries or datatype definitions.

An important technical novelty is a derivation of a weakened form of
positive-recursive type in Coq.  Coq (Agda, and Lean) restrict
datatypes $D$ to be strictly positive: in the type for any constructor
of $D$, $D$ cannot occur to the left of any arrows.  Our derivation
needs to use positive-recursive types, where $D$ may occur to the left
of an even number (only) of arrows.  We present a way to derive a
weakened form of positive-recursive type that is sufficient for our
examples (Section~\ref{sec:mu}).  The weakening is to require only
that $F\ \mu$ is a retract of $\mu$, where $\mu$ is the recursive type
and $F\ \mu$ its one-step unfolding.  Usually these types are
isomorphic.  Hence, we dub these \textbf{retractive-positive}
recursive types.  This weakening leads to noncanonical elements of
$\mu$, but we will see how to work around this.  Our definition of
retractive-positive recursive types makes essential use of impredicate
quantification, and hence is not available in predicative theories
like Agda's.

@@

We will use Coq inductive types for the signature functors \verb|F| of
various datatypes, thus enabling recursions to use Coq's
pattern-matching on the subdata structure \verb|d|.  So the style of
coding against this interface retains a similar feel to structural
recursion.  Unlike with structural termination, though, the interface
here is type-based and hence compositional.

As in previous work, we dub this interface a \emph{recursion
universe}~\cite{stump20}.  As in other domains using the term
``universe'', we have an entity (here, \verb|R|) from which one cannot
escape by using the available operations (for other cases:
the ordinal $\epsilon_0$ and $\omega^-$, the physical universe and traveling at
the speed of light).  Staying in the recursion universe is good,
because we may recurse (via \verb|eval|) on any value of type
\verb|R|.  Some points must still be explained: why \verb|X| has type \verb|Set -> Set|,
and the definition of \verb|FoldT|.  Let us see these details next.

@@

  This file specializes the development in \verb|Subrec.v| to the case
  of lists (parametrized by the type \verb|A| of elements).  In
  general, to use our development to get subsidiary recursion over
  some datatype, one will have a similar ``shim'' file.  For space
  reasons, we just give the example of lists.  The file defines the
  signature functor \verb|ListF|, shown in Figure~\ref{fig:listf}.  We
  then define \verb|List| to be \verb|Subrec|, with the instantiation
  of \verb|F| to \verb|ListF A|.  This type \verb|List| is not to be
  confused with the type \verb|list| of lists in Coq's standard
  library.  As noted previously, our development is meant to be used
  in extension of existing inductive datatypes, not replacing them.
  The figure also shows constructors \verb|mkNil| and \verb|mkCons|
  for \verb|List|, and types for conversion functions between
  \verb|List| and \verb|list|; one direction uses Coq's structural
  recursion, the other uses our subsidiary recursion (code elided).

@@

\verb|Subrec.v| also defines the notion of algebra that is used for
writing recursions.  The central definitions are in
Figure~\ref{fig:algf}.  \verb|KAlg| is the kind for the
type-constructor for algebras, as we see in the definition of
\verb|Alg|.  This type-constructor \verb|Alg| is a fixed-point of the
type \verb|AlgF|.  The fixed-point is taken using \verb|MuAlg|, which
implements our retractive-positive recursive types
(Section~\ref{sec:mu}) at kind \verb|KAlg|.  Using \verb|Alg| will
require that \verb|AlgF| only uses its parameter \verb|Alg|
positively.  We will confirm this shortly.

@@

\verb|FoldT Alg C| is the type for fold functions which apply
algebras of type \verb|Alg| to data of type \verb|C|, which we have
already dubbed the \emph{anchor type} of the recursion.  At the top
level of code, the anchor type would just be \verb|List| (for
example).  When one initiates a subsidiary recursion, though, the
anchor type will instead by the abstract type \verb|R| for the outer recursion.
The variable \verb|Alg| occurs only positively (but not strictly
positively) in \verb|AlgF|, because it occurs negatively in
\verb|FoldT Alg R| which occurs negatively in \verb|AlgF Alg C X|.  So
we can indeed take a fixed-point of \verb|AlgF| to define the constant
\verb|Alg|.

@@

Given a predicate \verb|p : A -> bool|, and a value of type
\verb|List A|, we would like to compute a pair of type
\verb|list A * List A|, where the first component is the maximal
prefix whose elements satisfy \verb|p|, and the second is the
remaining suffix.  This is the typing for a top-level recursion.  More
generally, though, given an anchor type \verb|R : Set| along with a
fold function for that anchor type (i.e., of type \verb|FoldT (Alg (ListF A)) R|),
we would like to map an input list of type \verb|R| to a pair of
type \verb|list A * R|.  The first component of this pair is
going to be built up from scratch, and so cannot have type \verb|R|;
we cannot statically ensure that outer recursions on it are legal.
But the second component will be a subdatum of the input list,
and so can still have type \verb|R|, enabling outer recursive calls.  So we want: 
\begin{verbatim}
Definition spanr{R : Set}(fo:FoldT (Alg (ListF A)) R)
                (p : A -> bool)(xs : R) : list A * R.
\end{verbatim}
\noindent From this we can also define the top-level recursion, by
supplying \verb|fold (ListF A)|, which is the function for folding an algebra
over a list (Figure~\ref{fig:algf}), for the argument \verb|fo| of \verb|spanr|:
\begin{verbatim}
Definition span(p : A -> bool)(xs : List A) : list A * List A
  := spanr (fold (ListF A)) p xs.
\end{verbatim}

Before we define \verb|spanr|, we must resolve a small problem.
If the first element of the input list
\verb|xs| to \verb|span| does not satisfy \verb|p|, then \verb|span|
should return \verb|([], xs)|.  But when recursing on \verb|xs|, we
will see it only in the form of a subdata structure of type
\verb|F R|.  We will not be able to return it from our recursion at
type \verb|R|, and hence we would not be able to return \verb|([],xs)|
as desired.  To work around this, we will have our recursion return a value
of type \verb|SpanF R| (\verb|X| will be implicit for the constructors):
\begin{verbatim}
Inductive SpanF(X : Set) : Set :=
  SpanNoMatch : SpanF X
| SpanSomeMatch : list A -> X -> SpanF X.
\end{verbatim}
\noindent The idea is that the recursion will signal if it is in the
one tricky case where \verb|p| does not match the first element, by
returning \verb|SpanNoMatch|.  Otherwise, it will be able to return,
via \verb|SpanSomeMatch|, a prefix and the suffix at type
\verb|R|. The prefix will be nonempty, and hence the suffix will be at
most the tail of \verb|xs|.  This tail is available to the algebra
in the subdata structure of type \verb|F R|.

\begin{figure}
\begin{verbatim}
Definition SpanAlg(p : A -> bool)(C : Set)
  : Alg (ListF A) C SpanF :=
  rollAlg (fun R reveal fo span xs => 
     match xs with
         Nil => SpanNoMatch 
       | Cons hd tl =>
          if p hd then
            match (span tl) with
              SpanNoMatch => SpanSomeMatch [hd] tl
            | SpanSomeMatch l r => SpanSomeMatch (hd::l) r
            end
          else
            SpanNoMatch 
       end).
\end{verbatim}
\caption{The algebra \texttt{SpanAlg} for the \texttt{span} function}
\label{fig:spanalg}
\end{figure}

Figure~\ref{fig:spanalg} gives the algebra \verb|SpanAlg| for computing
\verb|span|.  The type of \verb|SpanAlg p C| is
\begin{verbatim}
Alg (ListF A) C SpanF
\end{verbatim}
This states that we are defining an algebra (\verb|Alg|) for the
\verb|ListF A| functor, with anchor type \verb|C| and carrier
\verb|SpanF|.  \verb|SpanF| has type \verb|Set -> Set|, as required
for the carriers of our algebras. The definition of \verb|SpanAlg| is
actually parametrized by \verb|C|, which is good, as it means we can
use \verb|SpanAlg| for top-level or subsidiary recursions.

Let us continue through the code for \verb|SpanAlg|
(Figure~\ref{fig:span}).  We use \verb|rollAlg| to create an algebra
from something whose type is an application of \verb|AlgF|.  This
takes in all the components of the recursion universe: the abstract
type \verb|R|, the \verb|reveal| function (not needed in this case),
the fold function (\verb|fo|) for any subsidiary recursions (also not
needed here), a function we choose to name \verb|span| for making
recursive calls, and finally \verb|xs : ListF A R|.  The algebra
pattern-matches on this \verb|xs|.  In the cases where it is empty or
where its head (\verb|hd|) does not satisfy \verb|p|, we return
\verb|SpanNoMatch|.  This signals to the caller that we really wished
to return \verb|([],xs)|, but could not because we do not have
\verb|xs| at type \verb|R|.  If the head does satisfy \verb|p|, then
we recurse on the tail (\verb|tl : R|) by calling the provided \verb|span : R -> SpanF R|.
If \verb|span tl|
returns \verb|SpanNoMatch|, that means that we should make \verb|tl|
the suffix in the pair we return (via \verb|SpanSomeMatch|).  Happily,
we have \verb|tl : R| here, so we can do this.  In either case (for
return value of \verb|span tl|), we add the head to the front of the
prefix.

\verb|SpanAlg| is used in the definition of \verb|spanhr|, in
Figure~\ref{fig:span}.  This function invokes the fold function it is
given, on \verb|SpanAlg|.  The final twist is now in the definition of
\verb|spanr|.  We call \verb|spanhr| on the input \verb|xs : R|.  If
\verb|spanhr| returns \verb|SpanNoMatch|, then we are supposed to
return \verb|([],xs)|, which we can do here, because we have
\verb|xs : R|.  It was only inside the algebra that we lost the
information that the subdata structure of type \verb|F R| is derived
from a value of type \verb|R|.  If \verb|spanhr| returns
\verb|SpanSomeMatch|, then the return value gives us the nonempty
prefix (\verb|l|) and the suffix (\verb|r|), which we then return.  We
also define a version of \verb|break| for subsidiary recursion.

\begin{figure}
\begin{verbatim}
Definition spanhr{R : Set}(fo:FoldT (Alg (ListF A)) R)
                 (p : A -> bool)(xs : R) : SpanF R :=
  fo SpanF SpanFunctor (SpanAlg p R) xs.

Definition spanr{R : Set}(fo:FoldT (Alg (ListF A)) R)
                (p : A -> bool)(xs : R) : list A * R
  := match spanhr fo p xs with
       SpanNoMatch => ([],xs)
     | SpanSomeMatch l r => (l,r)
     end.

Definition breakr{R : Set}(fo:FoldT (Alg (ListF A)) R)
                 (p : A -> bool)(xs : R) : list A * R :=
  spanr fo (fun x => negb (p x)) xs.
\end{verbatim}
\caption{Functions derived from \texttt{SpanAlg}}
\label{fig:span}
\end{figure}

@@

Let us now see how to write \verb|wordsBy|, our example function from
Section~\ref{sec:intro}, using \verb|breakr| subsidiarily.
The code is in Figure~\ref{fig:wordsby}, assuming a type
\verb|A : Set|.  The setup is similar to that for \verb|span|.  We first define
an algebra \verb|WordsBy|, parametrized by anchor type \verb|C| (and also
the predicate \verb|p|), of type
\begin{verbatim}
Alg (ListF A) C (Const (list (list A)))
\end{verbatim}
\noindent This says that \verb|WordsBy p C| is an algebra (\verb|Alg|)
for the \verb|ListF A| functor, with anchor type \verb|C|, and carrier
\verb|Const (list (list A))|.  \verb|Const| is the combinator for
creating the object part of constant functors; \verb|FunConst| creates
the morphism part (i.e., the \verb|fmap| function).  We use 
\verb|Const| where the return type of the
algebra will not depend on its abstract type \verb|R|.  Here, we are
constructing from scratch a list of lists, so it will not be legal to
recurse on the list itself, or its (list) elements.  So we just
use the \verb|list| type of Coq's standard library.

@@

Here we see the benefit of our approach.  From Figure~\ref{fig:span},
the return type of \verb|breakr| is \verb|list A * R|, where \verb|R|
is the anchor type of the provided fold function \verb|fo|.  And
\verb|fo| has type \verb|FoldT (ListF A) Alg R|, from the definition
of \verb|AlgF| in Figure~\ref{fig:algf} (instantiating the functor
with \verb|ListF A|).  This means that from the invocation of
\verb|breakr|, we get \verb|w : list A| and \verb|z : R|.  And so we
can indeed apply \verb|wordsBy : R -> list (list A)| to
\verb|z| to recurse.

@@

\begin{figure}
\begin{verbatim}
mapThrough :: (a -> [a] -> (b, [a])) -> [a] -> [b]
mapThrough f [] = []
mapThrough f (a:as) = b : mapThrough f as'
    where (b, as') = f a as
\end{verbatim}
\caption{The \texttt{mapThrough} function in Haskell}
\label{fig:mapthroughhs}
\end{figure}

The Haskell library \verb|Data.List.Extra| has a function
\verb|repeatedly|, defined essentially as in
Figure~\ref{fig:mapthroughhs}. We attempt a more informative
name.  This is like the standard \verb|map| function on lists, except
that the function \verb|f| that we are mapping (or ``mapping
through'') takes in not just the current element \verb|a|, but also
the tail \verb|as|.  It then returns the value \verb|b| to include in
the output list, and whatever other list it wishes, upon which
\verb|mapThrough| will recurse.

To write this combinator using our infrastructure for subsidiary
recursion, we need to supply the mapped function with the fold
function for \verb|mapThrough|'s recursion.  This is so that the
mapped function can initiate a subsidiary recursion, returning a value
in the abstract type \verb|R| of \verb|mapThrough|'s recursion.  So
the type we will use for mapped functions is:
\begin{verbatim}
Definition mappedT(A B : Set) : Set :=
  forall(R : Set)(fo:FoldT (Alg (ListF A)) R), A -> R -> B * R.
\end{verbatim}
\noindent This type is more informative than the Haskell type,
since it shows that the second component of the returned value
must have type \verb|R|, and hence must be (hereditarily) a tail
of the input.

Given this definition, the code is in Figure~\ref{fig:mapthrough}.  
\verb|MapThroughAlg| is very similar (discounting syntax) to the
Haskell code above.  Here, though, when we call \verb|f|, we must
supply the abstract type \verb|R| and fold function \verb|fo|.  Then,
from the definition of \verb|mappedT|, we have that \verb|b : B| and
\verb|c : R|.  So we may indeed invoke \verb|mapThrough : R -> list B|
on \verb|c|.  Note that as we are building up a new list from scratch
(rather than just extracting some tail of the input list), we just
return \verb|list B|; we cannot perform further subsidiary recursion
on the output.

@@

Definition mapThroughr{R : Set}(fo:FoldT (Alg (ListF A)) R)
                      {B : Set}(f:mappedT A B) : R -> list B.
Definition mapThrough{B : Set}(f:mappedT A B) : List A -> list B.
\end{verbatim}
\caption{The algebra \texttt{MapThroughAlg} defining function \texttt{mapThrough} and \texttt{mapThroughr}; the
  code for those follows the pattern of \texttt{wordsBy} and \texttt{wordsByr} (Figure~\ref{fig:wordsby}), so
  we omit it}
\label{fig:mapthrough}
\end{figure}
